Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-06-06 paper.tex
Rearrange test section: narrative of hierarchy of physical realism

This is a challenging task in the environment of a merger,
which generally lacks any spatial symmetries so that fully general
seven-dimensional solutions to the Boltzmann Equation are not feasible.
Leakage approximations
\citep{deat2013-leakage, pere2016-asl, radi2016-dynamical, ?}
capture the qualitative effects of neutrinos on the matter, but
produce limited information about the neutrino field itself.
Monte Carlo methods \citep{abdi2012-monte_carlo, rich2015-monte_carlo}
suffer from noise.
\todo{discuss mc more clearly}
The state of the art today employs a truncated moment formalism
\citep{shib2011-truncated_moment, fouc2015-m1_nsbh, fouc2016-m1_nsns,
  just2015-m1_code, ?},
evolving the zeroth- and first-angular moments of the distribution function
(the energy density and flux respectively) representing the zeroth-energy moment
of the radiation field (the total energy).
Recently, \cite{fouc2016-m1_evolve_n} have expanded their code to also evolve
the number densities of neutrinos, providing an estimate of the average neutrino
energies in addition to their totals.
But with only two angular moments and one energy moment, the distribution
function has very limited angular and spectral information.

\subsubsection*{What physics problems do we want to address?}
But many interesting unsolved problems require an accurate model of the neutrino
spectra and angular distributions.
With a model of the neutrino emission from a merger we can
1) examine neutrino effects on the nucleosynthesis of the ejected material
\citep{surm2011-nickel_56, robe2016-sph_nu_nucleo},
2) explore the rich flavor oscillation physics likely to occur
\citep{malk2012-mnr_1, malk2015-mnr_2, malk2016-mnr_3, zhu2016-mnr_nsns_remnant,
  vaan2016-uncovering_mnr},
3) improve closure relations used in truncated moment schemes
\citep{ramp2002-truncated_moment, shib2011-truncated_moment,
  card2013-truncated_moment, fouc2015-m1_nsbh, ocon2015-gr1d_with_nu}, and
4) study possible jet formation due to neutrino annihilation
\citep{ruff1999-nunubar_nsns, asan2000-nunubar, birk2007-nunubar,
  hari2010-gr_nunubar_collapsar, zala2011-nunubar, leng2014-nunubar}.

@@

We build upon existing ray tracing formulations
by eschewing any assumptions about the spacetime geometry
and by including elastic scattering in the integration along each
geodesic.
To incorporate scattering processes in our method, we employ estimates of the
lowest-order moment of the neutrino distribution function
computed in a truncated moment scheme (formulated by
\cite{thor1981-truncated_moment, shib2011-truncated_moment}
and applied e.g. in \cite{ramp2002-truncated_moment,
  card2013-truncated_moment, fouc2015-m1_nsbh, ocon2015-gr1d_with_nu}).
If moments are not available either from
a truncated moment evolution or a trustworthy analytical estimate,
our method reduces to the current state-of-the-art ray tracing method,
neglecting scattering.

@@

This paper introduces the use of neutrino densities and fluxes evolved in a
truncated moment simulation to estimate the background field, $\Phi_{\rm SE}$.
App.~\ref{ssec:sources_se} details a method to calculate
$\Phi_{\rm SE}(\varepsilon)$ in two different ways:
\begin{itemize}
\item
  the \emph{spectral} method using densities and fluxes
  extracted from a simulation evolved over multiple energy groups
  to compute the background field with Eqn.~\ref{eqn:se_factor_expansion},
\item
  the \emph{gray} method using energy-integrated densities and fluxes
  extracted from a gray simulation and approximating the
  energy dependence with Eqns.~\ref{eqn:J_from_gray} and
  \ref{eqn:H_from_gray}.
\end{itemize}

@@

To test our algorithm, we integrate the ray tracing equations
(Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau}) in the following configurations,
and compare $f$ or moments of $f$ to known analytic solutions, or in some
cases to fiducial numerical calculations.

\subsection{Homogeneous Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter in two homogeneous configurations
representative of the fluid in the Massive Collapsed Star test
(Sec.~\ref{ssec:test_collapse}) at two different radii:
at 50~km
For comparision with Collapsed Star test, we use the LS180 equation of state,
in which the equilibrium
neutrino chemical potentials are
$\mu_{\nu_e}=-0.639$~MeV, and 
$\mu_{\nu_e}=5.771$~MeV respectively
($\mu_{\bar{\nu}_e}=-\mu_{\nu_e}$ and $\mu_{\nu_x}=0$).
Plot in Fig.~\ref{fig:homogeneous_isotropic}.
\todo{run test at this thermodynamic point}

\subsection{Doppler Shift}
\label{ssec:test_doppler}
We reproduce the test above, but with the matter moving in
in the $+z$ direction, with the observer stationary.
Plot in Fig.~\ref{fig:avg_eps_doppler}.
\todo{describe test}

\subsection{Gravitational Redshift and Geodesic Curvature}
\label{ssec:test_gravity}
Optically thick star in Schwarszchild.
\todo{describe $L(r)$, as a function of lapse}
optically thick disk in Kerr (Fishbone-Moncrief disk):
set up to give two different surfaces for nue vs nua, but same temp (?).
\todo{plot of analytic and numerical geodesics}
\todo{plot of $G$ nue vs nua}

\subsection{Homogenous Absorbing Star}
\label{ssec:test_ab_star}
[Test described in~\cite[Sec.~3.2]{smit1997-two_moment}.]
Plot in Fig.~\ref{fig:f_absorption_sphere}.

\subsection{Massive Collapsed Star with Scattering Envelope}
\label{ssec:test_collapse}
[Standard radiation transport test described in~\cite{ocon2015-gr1d_with_nu},
and in \cite[App.~E.6]{fouc2015-m1_nsbh},
and in~\cite{abdi2012-monte_carlo}.]
Plot in Fig.~\ref{fig:avg_eps_collapse}.

\subsection{Disk Comparison to M1}
\label{ssec:test_disk_comparison}
[Compute total and relative number and energy luminosities for a disk evolved
in one of \cite{fouc2015-m1_nsbh, fouc2016-m1_nsns, fouc2016-m1_evolve_n}.
Expect agreement within factors of a few (?).]
\todo{compute luminosities from M1 and RT, present table}

>>>>

This is a challenging task in the environment of a merger,
which generally lacks any spatial symmetries so that fully general
seven-dimensional solutions to the Boltzmann Equation are not feasible.
Leakage approximations
\citep{deat2013-leakage, pere2016-asl, radi2016-dynamical, ?}
capture the qualitative effects of neutrinos on the matter, but
produce limited information about the neutrino field itself.
Monte Carlo methods \citep{abdi2012-monte_carlo, rich2015-monte_carlo}
suffer from noise.
\todo{discuss mc more clearly}
The state of the art today couples radiation and matter with a truncated
moment formalism
\citep{shib2011-truncated_moment, fouc2015-m1_nsbh, fouc2016-m1_nsns,
  just2015-m1_code, ?},
evolving the zeroth- and first-angular moments of the distribution function
(the energy density and flux respectively) representing the zeroth-energy moment
of the radiation field (the total energy).
We refer to this as M1 transport.
Recently, \cite{fouc2016-m1_evolve_n} have expanded their code to also evolve
the number densities of neutrinos, providing an estimate of the average neutrino
energies in addition to their totals.
But with only two angular moments and one energy moment, the distribution
function has very limited angular and spectral information.

\subsubsection*{What physics problems do we want to address?}
But many interesting unsolved problems require an accurate model of the neutrino
spectra and angular distributions.
With a model of the neutrino emission from a merger we can
1) examine neutrino effects on the nucleosynthesis of the ejected material
\citep{surm2011-nickel_56, robe2016-sph_nu_nucleo},
2) explore the rich flavor oscillation physics likely to occur
\citep{malk2012-mnr_1, malk2015-mnr_2, malk2016-mnr_3, zhu2016-mnr_nsns_remnant,
  vaan2016-uncovering_mnr},
3) improve closure relations used in M1 transport schemes
\citep{ramp2002-truncated_moment, shib2011-truncated_moment,
  card2013-truncated_moment, fouc2015-m1_nsbh, ocon2015-gr1d_with_nu}, and
4) study possible jet formation due to neutrino annihilation
\citep{ruff1999-nunubar_nsns, asan2000-nunubar, birk2007-nunubar,
  hari2010-gr_nunubar_collapsar, zala2011-nunubar, leng2014-nunubar}.

@@

We build upon existing ray tracing formulations
by eschewing any assumptions about the spacetime geometry
and by including elastic scattering in the integration along each
geodesic.
To incorporate scattering processes in our method, we employ estimates of the
lowest-order moment of the neutrino distribution function
computed in an M1 transport simulation (formulated by
\cite{thor1981-truncated_moment, shib2011-truncated_moment}
and applied e.g. in \cite{ramp2002-truncated_moment,
  card2013-truncated_moment, fouc2015-m1_nsbh, ocon2015-gr1d_with_nu}).
If moments are not available either from
an M1 transport evolution or a trustworthy analytical estimate,
our method reduces to the current state-of-the-art ray tracing method,
neglecting scattering.

@@

This paper introduces the use of neutrino densities and fluxes evolved in an
M1 transport simulation to estimate the background field, $\Phi_{\rm SE}$.
App.~\ref{ssec:sources_se} details a method to calculate
$\Phi_{\rm SE}(\varepsilon)$ in two different ways:
\begin{itemize}
\item
  the \emph{spectral} method using densities and fluxes
  extracted from a simulation evolved over multiple energy groups
  to compute the background field with Eqn.~\ref{eqn:se_factor_expansion},
\item
  the \emph{gray} method using energy-integrated densities and fluxes
  extracted from a gray simulation and approximating the
  energy dependence with Eqns.~\ref{eqn:J_from_gray} and
  \ref{eqn:H_from_gray}.
\end{itemize}

@@

To test our algorithm, we integrate the ray tracing equations
(Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau}) at one observation position
in the following configurations. This suite of configurations
may be understood as a hierarchy of models of increasing
physical realism: beginning with a homogeneous medium of effectively
infinite extent (i.e.\ optically thick) and progressing to a model of a
1D pre-supernova, post-bounce collapse profile
evolved using an M1 transport hydrodynamical simulation.
\todo{also nsns merger?}

We compare to analytic solutions in the simple configurations, and
numerical solutions from alternative radiation transport algorithms
(i.e.\ monte carlo or M1 transport) where they are not.

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter in two homogeneous configurations
representative of the fluid in the Massive Collapsed Star test
(Sec.~\ref{ssec:test_collapse}) at two different radii:
at 50~km
For comparision with Collapsed Star test, we use the LS180 equation of state,
in which the equilibrium
neutrino chemical potentials are
$\mu_{\nu_e}=-0.639$~MeV, and 
$\mu_{\nu_e}=5.771$~MeV respectively
($\mu_{\bar{\nu}_e}=-\mu_{\nu_e}$ and $\mu_{\nu_x}=0$).
Plot in Fig.~\ref{fig:homogeneous_isotropic}.
\todo{run test at this thermodynamic point}

\subsection{Infinite Homogeneous Moving Slab:
  Testing Doppler Shift}
\label{ssec:test_doppler}
We reproduce the test above, but with the matter moving in
in the $+z$ direction, with the observer stationary.
Plot in Fig.~\ref{fig:avg_eps_doppler}.
\todo{describe test}

\subsection{Idealized Star:
  Testing the Decoupling Regime}
\label{ssec:test_ab_star}
[Test described in~\cite[Sec.~3.2]{smit1997-two_moment}.]
Plot in Fig.~\ref{fig:f_absorption_sphere}.

\subsection{Idealized Compact Star:
  Testing Gravitational Redshift and Geodesic Curvature}
\label{ssec:test_gravity}
Optically thick star in Schwarszchild.
\todo{describe $L(r)$, as a function of lapse}
optically thick disk in Kerr (Fishbone-Moncrief disk):
set up to give two different surfaces for nue vs nua, but same temp (?).
\todo{plot of analytic and numerical geodesics}
\todo{plot of $G$ nue vs nua}

\subsection{Post-Bounce Collapse Profile:
  Testing Scattering}
\label{ssec:test_collapse}
[Standard radiation transport test described in~\cite{ocon2015-gr1d_with_nu},
and in \cite[App.~E.6]{fouc2015-m1_nsbh},
and in~\cite{abdi2012-monte_carlo}.]
Plot in Fig.~\ref{fig:avg_eps_collapse}.

\subsection{Hypermassive Neutron Star Remnant:
  Testing Everything}
\label{ssec:test_disk_comparison}
[Compute total and relative number and energy luminosities for a disk evolved
in one of \cite{fouc2015-m1_nsbh, fouc2016-m1_nsns, fouc2016-m1_evolve_n}.
Expect agreement within factors of a few (?).]
\todo{compute luminosities from M1 and RT, present table}
